---
title: "Targeted exon sequencing - Automated QC"
date:  "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  pdf_document:
    latex_engine: tectonic
    keep_tex: true
---

```{r setup, include=FALSE}
options(tinytex.verbose = TRUE)
# load libraries
for(p in c("ggplot2", 'knitr', 'reshape2')){
  if(!p %in% installed.packages()[,1]){
    print(p)
    install.packages(p, repos =  "https://cloud.r-project.org", dependencies = T )
    library(p, character.only = TRUE)
  }
  library(p, character.only = TRUE)
}

# set global options
options(datatable.fread.datatable=FALSE)
options(stringsAsFactors = FALSE)

# TESTING ARGS
testing <- F
if(testing){
    # this pipeline actually just starts with sample reads
    sample.reads.f <- '~/data/project_one/sample_reads.tsv'
    # targets file with the probes
    targets.w0.f <- '~/data/references/AML_261_IDT_GRCh38/xgen_aml_probes.bed'
    # targets.old <- read.table('~/data/references/AML_261_IDT/targets_hg19.st.bed', sep='\t', header=F, quote='', col.names = c('chr', 'start', 'end', 'strand', 'exon'))
    # targets.old$length <- targets.old$end - targets.old$start
    outdir.server <- '/path/to/data/project_one/output'
    outdir <- '~/data/project_one/output'
} else {
    # set directories
    outdir <- snakemake@params[["outdir"]]
    # IDT probes file
    targets.w0.f <- snakemake@params[["target_bed"]]
    # sample reads file
    sample.reads.f <- snakemake@params[["sample_reads"]]
}
plot_output_dir <- file.path(outdir, 'QC_plots')
if (!dir.exists(plot_output_dir)){ dir.create(plot_output_dir) }

# read these files
sample.reads <- read.table(sample.reads.f, sep='\t', header=F, quote='', col.names = c('sample', 'r1', 'r2'))
targets.w0 <- read.table(targets.w0.f, sep='\t', header=F, quote='', col.names = c('chr', 'start', 'end', 'strand', 'exon'))
# set up 100 and 1000 windows
targets.w100 <- targets.w0
targets.w100$start <- targets.w100$start - 100
targets.w100$end <- targets.w100$end - 100
targets.w1000 <- targets.w0
targets.w1000$start <- targets.w1000$start - 1000
targets.w1000$end <- targets.w1000$end - 1000
targets.w0$length <- targets.w0$end - targets.w0$start
targets.w100$length <- targets.w100$end - targets.w100$start
targets.w1000$length <- targets.w1000$end - targets.w1000$start
# sample list
sample.list <- sample.reads$sample

```

## Introduction

This report is an automated check on the coverage of the targeted exons, for samples processed with a targeted sequencing panel, in all samples processed in the pipeline.

## Number of aligned reads
The first metric to study is the number of reads that can be uniquely aligned to the human genome in the exons of interest. We will calculate reads that align in the exons +/- 100bp. 

```{r, echo=F, message=F, warning=F, fig.width=12, fig.height=6}
# coverage info calculated from the pipeline
cov.dir <- file.path(outdir, '02_align/cov')
cov.w0.files <- list.files(cov.dir, pattern='bam.w0.cov$')
cov.w100.files <- list.files(cov.dir, pattern='bam.w100.cov$')

cov.w100.list <- lapply(cov.w100.files, function(x) {
    cov.w100 <- read.table(file.path(cov.dir, x), sep='\t', col.names = c('chr', 'start', 'end', 'strand', 'exon', 'reads', 'length', 'length.g0', 'pct.g0', 'avg.cov'))
    return(cov.w100)
})

sample.names <- sapply(cov.w100.files, function(x) strsplit(x, split='\\.')[[1]][1])
cov.w100.df <- cbind(targets.w100, do.call(cbind, lapply(cov.w100.list, function(x) x$avg.cov )))
cov.w100.n.df <- cbind(targets.w100, do.call(cbind, lapply(cov.w100.list, function(x) x$reads )))
colnames(cov.w100.df) <- c(colnames(targets.w100), sample.names)
colnames(cov.w100.n.df) <- c(colnames(targets.w100), sample.names)

# remove some rows with zero length from the GRCh37 to 38 liftover
cov.w100.df <- cov.w100.df[cov.w100.df$length >0, ]
nsamp <- length(cov.w100.files)

# total reads aligned per sample
total.reads.aligned <- data.frame(sample=sample.names, total.reads=colSums(cov.w100.n.df[,sample.names]))
total.reads.plot <- ggplot(total.reads.aligned, aes(y=sample,x=total.reads)) +
    geom_bar(stat='identity') + 
    theme_bw() + 
    labs(title='Count of reads aligned in exons +/- 100bp',x='# reads',y='Sample')
total.reads.plot

# calculate some basic stats
stat.df <- data.frame('Statistic' = c('total samples', 'mean read count', 'median', 'range'),
                      result = c(nrow(total.reads.aligned), 
                                 round(mean(total.reads.aligned$total.reads), 0),
                                 round(median(total.reads.aligned$total.reads), 0),
                                 paste(round(range(total.reads.aligned$total.reads), 0), collapse=' - ')))
kable(stat.df)
```
## Coverage
We will now look at coverage in the exons +/- 100bp. We will look for any samples that have particularly low coverage, as well as any exons that appear to be problematic across multiple samples. This first plot simply shows mean and median coverage across all exons. This can be used to identify areas with systematically low coverage (indicative of poor amplification) and areas of systematically high coverage (potentially indicative of artefacts).

```{r, echo=F, message=F, warning=F, fig.width=12, fig.height=6}
# look at stats across all exons
cov.w100.df$ALL_mean <- rowMeans(as.matrix(cov.w100.df[,sample.names]))
cov.w100.df$ALL_median <- apply(as.matrix(cov.w100.df[,sample.names]), 1, median)
cov.w100.df$ALL_max <- apply(as.matrix(cov.w100.df[,sample.names]), 1, max)
cov.w100.df$ALL_min <- apply(as.matrix(cov.w100.df[,sample.names]), 1, min)
cov.w100.df$n <- 1:nrow(cov.w100.df)

# annote with the chr breaks of the exon
chr.breaks.each <- sapply(unique(cov.w100.df$chr), function(x) sum(cov.w100.df$chr==x))
chr.breaks <- c(0, cumsum(sapply(unique(cov.w100.df$chr), function(x) sum(cov.w100.df$chr==x))))
chr.letters <- data.frame(letter=c(1:22, 'X'), pos=chr.breaks[1:23] + chr.breaks.each /2 )

toplot <- melt(cov.w100.df[, c('n', 'ALL_mean', 'ALL_median')], id.vars = 'n')
exon.genome.plot <- ggplot(toplot, aes(x=n, y=value)) +
    geom_line() +
    theme_bw() +
    geom_vline(xintercept = chr.breaks, col='firebrick') +
    geom_text(data=chr.letters, aes(x=pos, y=-700, label=letter), col='firebrick')+
    facet_wrap(.~variable, nrow = 2) + 
    labs(title='Coverage across all samples in all exons', x='Exon number', y='Coverage')
exon.genome.plot
ggsave(paste(format(Sys.time(), "%Y%m%d"), "coverage_exon_genome.pdf", sep="_"), 
       plot = exon.genome.plot,
       path = plot_output_dir,
       width =11, height = 8, units = c("in")) 
```
This first table shows the exons with <50x median coverage, which may be excluded from variant analysis. The second table highlights the 10 eoxns with the highest coverage.

```{r, echo=F, message=F, warning=F, fig.width=12, fig.height=6}
# what are the lowest mean locations?
# these are exons that are likely poor primers and under-covered in all the data
min.median.cov <- 50
low.cov <- cov.w100.df[cov.w100.df$ALL_median <= min.median.cov, ]
low.cov.report <- low.cov[order(low.cov$ALL_median), c('chr', 'start', 'end', 'strand', 'exon', 'length', 'ALL_median')]
kable(low.cov.report)

high.cov.report <- cov.w100.df[order(cov.w100.df$ALL_median,decreasing = T), c('chr', 'start', 'end', 'strand', 'exon', 'length', 'ALL_median')][1:10,]
kable(high.cov.report)
```
We will now look per-sample to see if there are any samples with coverage significantly lower than the rest. This boxplot shows coverage across all exons for each sample. If any samples are outliers, with mean coverage greater or less than the mean +/- two standard deviations, they will be printed below. 

```{r, echo=F, message=F, warning=F, fig.width=12, fig.height=6}
# coverage across all boxplot
cov.w100.boxplot.melt <- melt(cov.w100.df[,sample.names])
cov.w100.boxplot <- ggplot(cov.w100.boxplot.melt, aes(x=variable, y=value)) +
    geom_boxplot() +
    theme_bw()+
    coord_flip() +
    labs(title='Average coverage of each exon in each sample',
         y='Exon average coverage', x='sample')
cov.w100.boxplot
ggsave(paste(format(Sys.time(), "%Y%m%d"), "coverage_w100_boxplot.pdf", sep="_"), 
       plot = cov.w100.boxplot,
       path = plot_output_dir,
       width = 8, height = 11, units = c("in")) 
# 
# # if we ignore the problem exons highlighted above
# cov.w100.boxplot.melt.filt <- melt(cov.w100.df[!(cov.w100.df$exon %in% low.cov.report$exon), sample.names])
# ggplot(cov.w100.boxplot.melt.filt, aes(x=variable, y=value)) +
#     geom_boxplot() +
#     theme_bw()+
#     coord_flip() +
#     labs(title='Average coverage of each exon in each sample', subtitle = 'Filtered to remove problem exons highlighted above',
#          y='Exon average coverage', x='sample')

# do any samples fall 2sd or more below or above the mean
cm <- colMeans(cov.w100.df[,sample.names])
sample.cov.mean <- mean(cm)
sample.cov.sd <- sd(cm)
show.low.val <- sample.cov.mean - (2*sample.cov.sd)
show.high.val <- sample.cov.mean + (2*sample.cov.sd)
show.low.samps <- which(cm <= show.low.val)
show.high.samps <- which(cm >= show.high.val)

```
`r sum(cm <= show.low.val)` sample(s) that have less than the mean -2 sd coverage: `r show.low.samps`
`r sum(cm >= show.high.val)` sample(s) that have less than the mean -2 sd coverage: `r show.high.samps`

## Offtarget mapping
Next, we will calculate the number of reads that map to offtarget locations, greater than 1000bp from the targeted exons. We would like this to be a low fraction of total reads. 

```{r, echo=F, message=F, warning=F, fig.width=12, fig.height=6}
# do something with the flagstats offtarget files
off.dir <- file.path(outdir, '02_align/offtarget')
off.files <- list.files(off.dir, pattern='flagstat$')

# read these
off.list <- lapply(off.files, function(x) {
    # print(x)
    as.integer(read.table(file.path(off.dir,x), sep=' ', header=F, fill=T)[,1][2:14])
})
off.df <- do.call(cbind, off.list)
colnames(off.df) <- sample.names
rownames(off.df) <- c("total", "secondary", "supplementary", "duplicates", "mapped", "paired", "read1", "read2", "properly.paired", "itself.and.mate.mapped", "singletons", "mate.mapped.to.a.different.chr", "mate.mapped.to.a.different.chr.mapQ.5")

total.reads.aligned$offtarget <- off.df["paired", total.reads.aligned$sample]
total.reads.aligned$offtarget.frac <- total.reads.aligned$offtarget / total.reads.aligned$total.reads
total.reads.off.plot <- melt(total.reads.aligned)
offtarget.barplot <- ggplot(total.reads.off.plot[total.reads.off.plot$variable %in% c('total.reads', 'offtarget'),],
       aes(x=value, y=sample, fill=variable)) +
    geom_bar(stat='identity') + 
    scale_fill_manual(name='', 
                      labels=c('On-target reads', 'Off-target reads'), 
                      values = c('steelblue', 'firebrick')) + 
    theme_bw() +
    labs(title='On-target vs off-target reads',
         x='# reads',y='Sample')
offtarget.barplot
ggsave(paste(format(Sys.time(), "%Y%m%d"), "offtarget_read_barplot.pdf", sep="_"), 
       plot = offtarget.barplot,
       path = plot_output_dir,
       width = 8, height = 11, units = c("in")) 

```
