---
title: "`r params$project_name`"
subtitle: "Somatic variant calling report"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
params:
  project_name: "Somatic variant analysis"
  outdir: "output"
  metadata: ""
  scripts_dir: "."
  genome_version: "GRCh38"
  filter_min_af: 0.02
  filter_min_dp: 5
  filter_min_cadd: 20
  filter_gatk: "PASS"
  top_n: 40
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    self_contained: true
    theme: flatly
    df_print: kable
  pdf_document:
    toc: true
    toc_depth: 2
    latex_engine: tectonic
    keep_tex: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 9, fig.height = 4.5, dpi = 150)
suppressPackageStartupMessages({
    library(ggplot2)
    library(data.table)
    library(knitr)
})
source(file.path(params$scripts_dir, "report_helpers.R"))

outdir <- params$outdir
ref_dir    <- file.path(outdir, "02_variants_reference")
annot_dir  <- file.path(ref_dir, "02_variant_annotations/annotations_combined/filtered")
called_dir <- file.path(ref_dir, "01_gatk_variant_calling/filtered")
scores_dir <- file.path(ref_dir, "03_variant_new_scores/final/filtered")
tvn_dir    <- file.path(outdir, "03_variants_TvN/01_gatk_variant_calling/filtered")
contam_dir <- file.path(outdir, "01_prepare_bam/CalculateContamination")
depth_file <- file.path(outdir, "01_prepare_bam/depth_smoothed_join.tsv")
facets_dir <- file.path(outdir, "04_FACETS")

variants  <- load_annotated_variants(annot_dir)
scores    <- load_new_scores(scores_dir)
vcf_counts<- load_vcf_counts(called_dir)
tvn_counts<- load_vcf_counts(tvn_dir)
contam    <- load_contamination(contam_dir)
depth     <- load_depth_summary(depth_file)
purity    <- load_facets_purity(facets_dir)

metadata <- NULL
if (nzchar(params$metadata) && file.exists(params$metadata)) {
    metadata <- tryCatch(
        data.table::fread(params$metadata, sep = "\t", showProgress = FALSE),
        error = function(e) NULL)
}

annotated_mode <- !is.null(variants)

# Summary tables are also written to disk so they can be opened in Excel / re-used
# downstream without re-running anything.
summary_dir <- file.path(outdir, "09_report", "summary_tables")
dir.create(summary_dir, recursive = TRUE, showWarnings = FALSE)
write_summary <- function(dt, name) {
    if (is.null(dt) || nrow(dt) == 0) return(invisible(NULL))
    data.table::fwrite(dt, file.path(summary_dir, paste0(name, ".tsv")), sep = "\t")
}
```

# Overview

```{r overview}
n_samples <- if (!is.null(metadata) && "sample" %in% names(metadata)) {
    length(unique(metadata$sample))
} else if (!is.null(vcf_counts)) nrow(vcf_counts) else NA_integer_

n_patients <- if (!is.null(metadata) && "patient" %in% names(metadata)) {
    length(unique(metadata$patient))
} else NA_integer_

n_tumor <- n_normal <- NA_integer_
if (!is.null(metadata) && "tumor_normal" %in% names(metadata)) {
    n_tumor  <- sum(metadata$tumor_normal %in% c("T", "tumor", "Tumor"))
    n_normal <- sum(metadata$tumor_normal %in% c("N", "normal", "Normal"))
}

overview <- data.table(
    Item = c("Samples analysed", "Patients", "Tumour samples", "Normal samples",
             "Reference build", "Variant annotation",
             "Min allele fraction", "Min depth", "Min CADD"),
    Value = c(as.character(n_samples), as.character(n_patients),
              as.character(n_tumor), as.character(n_normal),
              params$genome_version,
              if (annotated_mode) "CADD + Funcotator" else "not run (calling only)",
              as.character(params$filter_min_af), as.character(params$filter_min_dp),
              as.character(params$filter_min_cadd))
)
overview <- overview[!is.na(Value) & Value != "NA"]
kable(overview, caption = "Run configuration")
write_summary(overview, "run_overview")
```

Variants were called with GATK Mutect2 against the reference genome for every sample, and
additionally in matched tumour-vs-normal mode wherever a patient/timepoint had both a
tumour and a normal sample. Calls shown below have GATK's own filters applied; the
"filtered" set additionally requires the allele-fraction, depth
`r if (annotated_mode) "and CADD" else ""` thresholds listed above.

`r if (!annotated_mode) "**Note:** this run was executed without the annotation stage, so gene-level, functional-consequence, and deleteriousness sections are omitted. Variant counts and QC below are still complete."`

# Sequencing quality control

## Coverage depth

```{r depth}
if (is.null(depth)) {
    note_missing("Per-sample depth")
} else {
    write_summary(depth, "depth_per_sample")
    kable(depth[order(mean_depth)],
          digits = 1,
          col.names = c("Sample", "Mean depth", "Median depth", "Fraction of targets <50x"),
          caption = "On-target sequencing depth per sample")
}
```

```{r depth-plot, fig.height=3.6}
if (!is.null(depth)) {
    ggplot(depth, aes(x = reorder(sample, mean_depth), y = mean_depth)) +
        geom_col(fill = REPORT_PALETTE[1]) +
        coord_flip() +
        labs(title = "Mean on-target depth by sample", x = NULL, y = "Mean depth (x)") +
        report_theme()
}
```

`r if (!is.null(depth)) "Samples with low mean depth, or a high fraction of targets below 50x, will have reduced sensitivity for low-allele-fraction variants. Interpret an absence of calls in those samples with caution."`

## Tumour-in-normal contamination

```{r contamination}
if (is.null(contam)) {
    note_missing("Contamination estimates (requires matched normals)")
} else {
    write_summary(contam, "contamination")
    kable(contam[order(-contamination)], digits = 4,
          col.names = c("Sample", "Contamination", "Std. error"),
          caption = "GATK CalculateContamination estimates")
}
```

`r if (!is.null(contam)) "Contamination estimates above roughly 0.05 suggest tumour content in the matched normal (or cross-sample contamination), which can cause true somatic variants to be filtered out as germline. These are already used by FilterMutectCalls."`

# Variant summary

## Variants per sample

```{r variant-counts}
counts <- NULL
if (annotated_mode) {
    counts <- variants[, .(n_variants = .N), by = sample]
    if ("FILTER" %in% names(variants)) {
        pass <- variants[FILTER == "PASS", .(n_pass = .N), by = sample]
        counts <- merge(counts, pass, by = "sample", all.x = TRUE)
        counts[is.na(n_pass), n_pass := 0L]
    }
} else if (!is.null(vcf_counts)) {
    counts <- vcf_counts
}

if (is.null(counts)) {
    note_missing("Variant calls")
} else {
    write_summary(counts, "variants_per_sample")
    kable(counts[order(-n_variants)], caption = "Variant counts per sample")
}
```

```{r variant-counts-plot, fig.height=3.6}
if (!is.null(counts)) {
    ggplot(counts, aes(x = reorder(sample, n_variants), y = n_variants)) +
        geom_col(fill = REPORT_PALETTE[3]) +
        coord_flip() +
        labs(title = "Variants passing filters, by sample", x = NULL, y = "Variants") +
        report_theme()
}
```

## Matched tumour-vs-normal calls

```{r tvn}
if (is.null(tvn_counts)) {
    note_missing("Paired tumour-vs-normal calls")
} else {
    write_summary(tvn_counts, "variants_per_patient_TvN")
    kable(tvn_counts[order(-n_variants)],
          col.names = c("Patient / timepoint", "Variants", "PASS"),
          caption = "Paired tumour-vs-normal variant counts")
}
```

## Functional consequence

```{r consequence, fig.height=4.5}
if (!annotated_mode || !"variantClassification" %in% names(variants)) {
    note_missing("Functional consequence annotation")
} else {
    cls <- variants[!is.na(variantClassification) & variantClassification != "",
                    .(n = .N), by = variantClassification][order(-n)]
    write_summary(cls, "variant_classification")
    top_cls <- head(cls, 12)
    print(
        ggplot(top_cls, aes(x = reorder(variantClassification, n), y = n)) +
            geom_col(fill = REPORT_PALETTE[2]) +
            coord_flip() +
            labs(title = "Variant classification (all samples)", x = NULL, y = "Variants") +
            report_theme()
    )
    kable(cls, col.names = c("Classification", "Variants"),
          caption = "Variant classifications observed")
}
```

## Allele fraction and depth

```{r vaf, fig.height=4}
if (!annotated_mode || !"AF" %in% names(variants) || all(is.na(variants$AF))) {
    note_missing("Per-variant allele fractions")
} else {
    print(
        ggplot(variants[is.finite(AF)], aes(x = AF)) +
            geom_histogram(bins = 50, fill = REPORT_PALETTE[1], colour = NA) +
            labs(title = "Variant allele fraction distribution",
                 x = "Allele fraction", y = "Variants") +
            report_theme()
    )
}
```

```{r vaf-by-sample, fig.height=4}
if (annotated_mode && "AF" %in% names(variants) && any(is.finite(variants$AF))) {
    d <- variants[is.finite(AF)]
    if (length(unique(d$sample)) > 1 && length(unique(d$sample)) <= 40) {
        print(
            ggplot(d, aes(x = reorder(sample, AF, FUN = stats::median), y = AF)) +
                geom_boxplot(outlier.size = 0.4, fill = REPORT_PALETTE[5], alpha = 0.7) +
                coord_flip() +
                labs(title = "Allele fraction by sample", x = NULL, y = "Allele fraction") +
                report_theme()
        )
    }
}
```

## Predicted deleteriousness

```{r cadd, fig.height=3.6}
if (!annotated_mode || !"CADD_phred" %in% names(variants) ||
    all(is.na(variants$CADD_phred))) {
    note_missing("CADD deleteriousness scores")
} else {
    print(
        ggplot(variants[is.finite(CADD_phred)], aes(x = CADD_phred)) +
            geom_histogram(bins = 40, fill = REPORT_PALETTE[4], colour = NA) +
            geom_vline(xintercept = as.numeric(params$filter_min_cadd),
                       linetype = "dashed", colour = "grey30") +
            labs(title = "CADD phred score distribution",
                 subtitle = paste0("dashed line = reporting threshold (",
                                   params$filter_min_cadd, ")"),
                 x = "CADD phred", y = "Variants") +
            report_theme()
    )
}
```

# Recurrently mutated genes

```{r recurrent, fig.height=4.5}
if (!annotated_mode || !"hugoSymbol" %in% names(variants)) {
    note_missing("Gene-level annotation")
} else {
    genes <- variants[!is.na(hugoSymbol) & hugoSymbol != "" & hugoSymbol != "Unknown",
                      .(n_variants = .N, n_samples = uniqueN(sample)), by = hugoSymbol]
    genes <- genes[order(-n_samples, -n_variants)]
    write_summary(genes, "recurrent_genes")
    top_genes <- head(genes, 20)
    if (nrow(top_genes) > 0) {
        print(
            ggplot(top_genes, aes(x = reorder(hugoSymbol, n_samples), y = n_samples)) +
                geom_col(fill = REPORT_PALETTE[6]) +
                coord_flip() +
                labs(title = "Genes mutated in the most samples",
                     x = NULL, y = "Samples with a variant") +
                report_theme()
        )
        kable(top_genes, col.names = c("Gene", "Variants", "Samples"),
              caption = "Most recurrently mutated genes")
    }
}
```

# Prioritised variants

```{r top-variants}
top_tbl <- NULL
if (!is.null(scores)) {
    score_col <- intersect(c("score.new", "score_new"), names(scores))
    if (length(score_col) == 1) {
        setnames(scores, score_col, "score_new")
        scores[, score_new := suppressWarnings(as.numeric(score_new))]
        keep <- intersect(c("patient", "hugoSymbol", "proteinChange",
                            "variantClassification", "CHROM", "POS", "REF", "ALT",
                            "CADD_phred", "mean.T.nna", "mean.N.nna", "score_new"),
                          names(scores))
        top_tbl <- scores[order(-score_new), ..keep]
        top_tbl <- head(top_tbl, as.integer(params$top_n))
    }
} else if (annotated_mode && "CADD_phred" %in% names(variants)) {
    keep <- intersect(c("sample", "hugoSymbol", "proteinChange", "variantClassification",
                        "CHROM", "POS", "REF", "ALT", "CADD_phred", "AF", "DP"),
                      names(variants))
    top_tbl <- variants[order(-CADD_phred), ..keep]
    top_tbl <- head(top_tbl, as.integer(params$top_n))
}

if (is.null(top_tbl) || nrow(top_tbl) == 0) {
    note_missing("Ranked variant list")
} else {
    write_summary(top_tbl, "top_variants")
    kable(top_tbl, digits = 3,
          caption = paste("Top", nrow(top_tbl), "variants by priority score"))
}
```

`r if (!is.null(scores)) "Variants are ranked by the pipeline's composite score (deleteriousness x mean tumour allele fraction x tumour/normal log2 fold-change), which brings likely-real somatic events to the top. See the somatic manual for the exact definition." else if (!is.null(top_tbl)) "No matched normals were available, so variants are ranked by predicted deleteriousness alone."`

# Copy number and loss of heterozygosity

```{r facets}
if (is.null(purity)) {
    note_missing("FACETS copy-number output")
} else {
    write_summary(purity, "facets_purity")
    kable(purity[, .(patient_tp, value)],
          col.names = c("Patient / timepoint", "FACETS output"),
          caption = "FACETS purity/ploidy estimates")
    cat("\nPer-sample copy-number and LOH plots are in `04_FACETS/` as one PDF per patient-timepoint.\n")
}
```

# Methods and provenance

```{r provenance}
prov <- data.table(
    Item = c("Report generated", "Reference build", "Output directory",
             "Variant caller", "Annotation", "Copy number"),
    Value = c(format(Sys.time(), "%Y-%m-%d %H:%M:%S"),
              params$genome_version,
              normalizePath(outdir, mustWork = FALSE),
              "GATK Mutect2 (tumour-vs-reference and matched tumour-vs-normal)",
              if (annotated_mode) "CADD + GATK Funcotator" else "not run",
              if (!is.null(purity)) "FACETS" else "not run")
)
kable(prov, caption = "Provenance")
```

Full result tables backing every figure in this report are written as tab-separated files
to `09_report/summary_tables/`, so they can be opened directly in a spreadsheet or re-used
in downstream analysis.

**Interpretation caveats.** Somatic calls made against the reference alone (without a
matched normal) will include private germline variants; the gnomAD germline resource and
the tumour/normal comparison mitigate but do not eliminate this. Variant calls in regions
with low depth are unreliable regardless of the filters applied. Predicted deleteriousness
scores are computational estimates, not clinical determinations, and no variant here has
been clinically validated.
